---
title: "Final Project"
author: "Mandi Acevedo, Kevin Velasco, Bela Walkin"
format: html
editor: visual
---

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
library(tidyr)
library(tidymodels)
library(recipes)

```

# Data cleaning

## Download data

```{r, warning=FALSE, message=FALSE}
df <- read_csv("data/countypres_2000-2020.csv")

```

## Pivot wider

## Create Democrat/Republican margin for each county for each year

```{r}
clean_data <- df |> 
  filter(mode == "TOTAL") |>
  filter(year >= 2016) |>
  filter(party %in% c("DEMOCRAT", "REPUBLICAN")) |>
  select(-candidate) |>
  pivot_wider(names_from = party, values_from = candidatevotes) |>
  mutate(margin = DEMOCRAT - REPUBLICAN,
         winner = case_when(
           margin > 0 ~ "DEMOCRAT",
           margin < 0 ~ "REPUBLICAN"))
  
```

## Add additional predictors

# Set up testing environment

## Initial split

```{r}
set.seed(12071999)

modeling_sample <- initial_split(clean_data)

train <- training(modeling_sample)
test <- testing(modeling_sample)

```

## Exploratory analysis

## V-fold cross-validation

```{r, warning=FALSE, message=FALSE}

train_folds <- vfold_cv(data = train, v = 10)

```

# Testing models

## Create a recipe

```{r}
recipe <-
  recipe(formula = margin ~ ., data = train) |>
  step_naomit(all_predictors(), -all_outcomes()) |>
  step_dummy(all_nominal(), -all_outcomes())

```

## LASSO

## Decision Tree

## KNN

(Insert model here) has the lowest out of sample error rate

# Final model estimation on 2024 data

